---
title: "Are p-Values Uniformly Distributed Under the Null?"
output: html_document
author: "Peter E. Freeman - Carnegie Mellon University"
---

The setting: you wish to determine if the $p$-values that arise from
the chi-square goodness of fit test and from multinomial simulations
are uniformly distributed, which you would expect them to be if the null
hypothesis is correct.

```{r}
k <- 30
m <- 6
p <- rep(1/m,m)
set.seed(101)

num.rep <- 1000
p.chi   <- rep(NA,num.rep)
p.mult  <- rep(NA,num.rep)
num.sim <- 1000

for ( ii in 1:num.rep ) {
  X.obs       <- rmultinom(1,k,p)
  W           <- sum((X.obs-k*p)^2/(k*p))
  p.chi[ii]   <- 1 - pchisq(W,m-1)
  obs         <- dmultinom(X.obs,prob=p)
  X           <- rmultinom(num.sim,k,p)
  a           <- apply(X,2,function(x,p){dmultinom(x,prob=p)},p=p)
  p.mult[ii]  <- sum(a<=obs)/num.sim
}

df <- data.frame(p.chi,p.mult)
```

To determine if the $p$-values arising from repeated applications of
the chi-square goodness of fit test are uniformly distributed, we apply
the Kolmogorov-Smirnov test.

```{r fig.align='center',fig.width=4,fig.height=4}
library(ggplot2)

ggplot(data=df,mapping=aes(x=p.chi,y=after_stat(density))) +
  geom_histogram(fill="dodgerblue",bins=25) +
  geom_hline(yintercept=1,col="firebrick1",lty=2) +
  xlab("p-Value - Chi-Square Goodness of Fit Test")

cat("KS test p-value:",suppressWarnings(ks.test(p.chi,"punif"))$p.value,"\n")
```

```{r fig.align='center',fig.width=4,fig.height=4}
library(ggplot2)

ggplot(data=df,mapping=aes(x=p.mult,y=after_stat(density))) +
  geom_histogram(fill="dodgerblue",bins=25) +
  geom_hline(yintercept=1,col="firebrick1",lty=2) +
  xlab("p-Value - Multinomial Simulations")

cat("KS test p-value:",suppressWarnings(ks.test(p.mult,"punif"))$p.value,"\n")
```

We observe that in the low-count regime, where the approximations underpinning 
the chi-square goodness-of-fit test rapidly break down, we reject the null
hypothesis that the $p$-values from the chi-square goodness-of-fit test are
uniformly distributed, while for the multinomial simulations, we fail to reject
the null.
